
```{r,echo=FALSE}
## Set default options for the knitr RMD processing
knitr::opts_chunk$set(echo=FALSE,warning=FALSE,message=FALSE,fig.width=5,fig.height=5,cache=FALSE,autodep=TRUE, results="hide")
```

[//]: # (Library and data loading                                     )
```{r libraries,cache=FALSE}

source( "functions/functions.R" )
source( "functions/load_libraries.R" )
source( "functions/load_data.R" )
source( "functions/make_strata.R" )
```

```{r,echo=FALSE}
## Tables and figures numbering
## counters for manuscript. These are used to automatically number the
## figures, tables etc. in the manuscript. To change the order, modify the
## calls below.

## figures
fig_register(c("consort", "cv", "metabo", "signature", "profiles", "diff_res"),
             prefix="fig_")

## tables
#fig_register(c("samples", "diff_res"), prefix="tab_")
fig_register(c("diff_res"), prefix="tab_")

## suplementary tables
fig_register(c("recruitment", "criteria", "approvals", "tst", "sample_types", 
  "kfold", "sitecross", "samplecross", "topcompounds", 
  "unblinding", "unblinding_bysite", "metabo_performance", "metabo_red", "tbord_performance", #"tbord_rev_performance", 
  "diff_res", "timecourse", "lmer_enr",
  "sessionInfo"), prefix="s_tab_")

## suplementary figures
fig_register(c("scheme", "validation_scheme", "cv", "cross_cohort", # "cross_stype", 
             "pred_rev", "nvar",
             # "fd", 
             # "profiles",
             "profiles_mod", "smoking"), prefix="s_fig_")
```


[//]: # (Machine learning  -----)
```{r ml}
## mapping all IDs in the universe
group <- set$group
names(group) <- rownames(set)
group <- c(group, setNames(r.metabo$targets$group, rownames(r.metabo$targets)))
```

[//]: # (Machine learning: k-fold)
```{r mlkfold}
## this is a slow procedure. The k-fold is on donors (not samples) to
## avoid splitting the samples from the same donor between the training and
## test sets, which might result in overfitting.
## The results are cached since they take quite long to compute; to
## recompute, remove the file static_cache/ss3_kfold.rda
if(file.exists("static_cache/ss3_kfold.rda")) {
  load("static_cache/ss3_kfold.rda")
} else {
  set.seed(20170120)
  ss3_kfold <- sapply(ss3, 
    function(x) mykfold(x, 
                        response=set[colnames(x), "group"], 
                        grouping_var=set[colnames(x), "DONOR_ID"]), 
    simplify=F)
  save(ss3_kfold, file="static_cache/ss3_kfold.rda")
}
ss3_kfold.tab <- rocseriestable(ss3_kfold)
ss3_kfold.tab$Nvar <- sapply(ss3, nrow)
```

```{r mlsamplecross}
## cross sample type predictions.
## The results are cached since they take quite long to compute; to
## recompute, remove the file static_cache/stype_cross.rda
if(file.exists("static_cache/stype_cross.rda")) {
  load("static_cache/stype_cross.rda")
} else {
  set.seed(20170203)
  strata <- names(stype.ss3)
    
  stype.cross <- sapply(strata, function(xn) { 
    x <- stype.ss3[[xn]]

    # cross-predictions only
    subset <- strata[ !grepl(gsub("\\..*", "", xn), strata) ]
    sapply(subset, function(yn) {
      y <- stype.ss3[[yn]]
      xpredict(x=x, x.response=set123[colnames(x), "group"], y=y, y.response=set[ colnames(y), "group" ])
    }, simplify=FALSE)
  }, simplify=FALSE)
  save(stype.cross, file="static_cache/stype_cross.rda")
}
```

```{r mlsitecross}
## cross site predictions.
## The results are cached since they take quite long to compute; to
## recompute, remove the file static_cache/site_cross.rda
if(file.exists("static_cache/site_cross.rda")) {
  load("static_cache/site_cross.rda")
} else {

  set.seed(20170201)

  # all strata (cohort x time point), but not "TOT"
  strata <- names(ss3)[ !grepl("TOT", names(ss3)) ]
  strata <- strata[ grep("\\.all$", strata) ]

  # cross contains each-against-each CV of the strata
  # e.g. cross[[1]] are predictions of "SUN.all" on all strata
  site.cross <- sapply(strata, function(xn) {
    x <- ss3[[xn]]
    subset <- strata[ !grepl(gsub("\\..*", "", xn), strata) ]
    sapply(subset, function(yn) {
      y <- ss3[[yn]]
      xpredict(x=x, x.response=set123[colnames(x), "group"], y=y, y.response=set[ colnames(y), "group" ])
    }, simplify=FALSE)
  }, simplify=FALSE)

  save(site.cross, file="static_cache/site_cross.rda")
}



```{r blinded}
## the blinded predictions have been generated once; the code is accessible
## (see readme.txt), but the predictions have been locked in a static file
## (data/set4_predictions_weiner.csv),
## transmitted to the GC6 consortium and not
## re-calculated again. Here, they are only evaluated.
## The object "pred" is loaded by functions/load_data.R

set4 <- set[ rownames(pred), ]
pred.models <- colnames(pred)
pred.full <- sapply(pred.models, function(cn) {
    df <- data.frame(reality=set4$group, prediction=pred[,cn], decision=pred.decision[,cn], row.names=rownames(set4))
    list(all=df, proximate=df[ set4$TIME_TO_TB < cutoff, ], distal=df[ set4$TIME_TO_TB >= cutoff, ], BL=df[ set4$TIMEPOINT == "BL", ])
}, simplify=FALSE)
pred.full <- unlist(pred.full, recursive=FALSE)
pred.full.tab <- rocseriestable(pred.full)


pred.bysite <- sapply(c("TotalF", "Total.BL"), function(cn) {
    df <- data.frame(reality=set4$group, prediction=pred[,cn], decision=pred.decision[,cn])
    sapply(sites, function(s) df[ set4$SITE == s, ], simplify=FALSE)
}, simplify=FALSE)
pred.bysite <- unlist(pred.bysite, recursive=FALSE)
pred.bysite.tab <- rocseriestable(pred.bysite)


pred.bl <- sapply(c("TotalF", "Total.BL"), function(cn) {
    df <- data.frame(reality=set4$group, prediction=pred[,cn], decision=pred.decision[,cn])
    sapply(sites, function(s) df[ set4$SITE == s & set4$TIMEPOINT == "BL", ], simplify=FALSE)
}, simplify=FALSE)
pred.bl <- unlist(pred.bl, recursive=FALSE)
pred.bl.tab <- rocseriestable(pred.bl)
```

```{r reduced_model_2}
## the reason that we are creating a new reduced model is that the original
## model by Fergal Duffy included unknown variables and did not include
## information on variable importance, necessary for a resulting list
## Note that this code was written after unblinding of the data, however it
## still uses the validation set to determine the model performance.
# set.seed(20170207)
# sel <- !grepl("^X - ", rg$genes$BIOCHEMICAL)
# reduced.model <- xpredict(x=x[sel,], y=y, x.response=set123$group, y.response=set4$group, nvar=15, return.list=TRUE)
# pred.reduced <- list(reduced.all=reduced.model$ret)
# 
# ssel <- set4$TIME_TO_TB < cutoff
# pred.reduced$reduced.proximate <- xpredict(x=x[sel,], y=y[ , ssel ], x.response=set123$group, y.response=set4$group[ssel], rf=reduced.model$rf)
# pred.reduced$reduced.distal    <- xpredict(x=x[sel,], y=y[ , !ssel], x.response=set123$group, y.response=set4$group[!ssel], rf=reduced.model$rf)
# pred.reduced.tab <- rocseriestable(pred.reduced)
```

```{r pred_enrichment}
## we test the models for enrichment
## models are stored in the data directory
## goal is to see which classes of substances are important for the models
require(tmod)
data(modmetabo)
## we will only use modules for which there are at least 5 metabolites in
## our data set

load("data/rf.total.rda")
load("data/rf.total.bl.rda")

m.l <- sapply(modmetabo$MODULES2GENES, function(s) sum(s %in% all.biochemicals[ rownames(rf.total$importance), ]$HMDB))
m.l <- sapply(modmetabo$MODULES2GENES, function(s) sum(s %in% all.biochemicals$HMDB))
m.l2 <- lengths(modmetabo$MODULES2GENES)
mm <- modmetabo[ m.l > 5 & m.l2 > 5 ]

res <- list()
g <- all.biochemicals[ rownames(rf.total$importance)[ order(rf.total$importance[,4], decreasing=T) ], ]$HMDB
res$Total <- tmodUtest(g, mset=mm, qval=Inf)
g <- all.biochemicals[ rownames(rf.total$importance)[ order(rf.total.bl$importance[,4], decreasing=T) ], ]$HMDB
res[["Total / Baseline"]] <- tmodUtest(g, mset=mm, qval=Inf)

## we only consider positive enrichment
res <- sapply(res, function(x) x <- x[ x$AUC > 0.5, ], simplify=FALSE)
res <- sapply(res, function(x) { x$adj.P.Val <- p.adjust(x$P.Value, method="fdr") ; x }, simplify=FALSE)

res.enrichment.rfs <- res
rm(res)
```

```{r simula}
### we use sampling to determine the optimal number of features in a model
### based on the metabo data set
if(file.exists("static_cache/simula.rda")) {
  load("static_cache/simula.rda")
} else {
  set.seed(20180807)
  sel <- Reduce(intersect, c(list(r.metabo$genes$ID), sapply(ss4, rownames)))
  response <- factor(r.metabo$targets$group)

  steps <- c(3:11, 13, 15, seq(20, 100, by=10), 150, 200)
  n <- 100

  res.sim <- sapply(steps, function(s) {
    message(s)
    replicate(n, {
        ret <- mykfold2(r.metabo$E[sel,], response, var=s)
        auc(ret)[2] })
    })

  colnames(res.sim) <- paste0("N.", steps)
  save(res.sim, file="static_cache/simula.rda")
}

```

```{r mlmetabo}
## The Metabo data set is directly derived from Weiner et al. 2012
## That is, it is a completely independent data set. We use it here to test
## the efficiency of a biosignature profiled on the disease in discriminating the
## early stages of the disease.
## Here: cross-predictions of metabo on *all* of GC6
set.seed(20170110)
sel <- Reduce(intersect, c(list(r.metabo$genes$ID), sapply(ss4, rownames)))
response <- factor(r.metabo$targets$group)
mod.metabo.full <- randomForest(t(r.metabo$E[sel, ]), response, importance=TRUE)

## construction of the shrinked model based only on known compounds
sel <- intersect(sel, r.metabo$genes$ID[ !grepl("^X - ", r.metabo$genes$BIOCHEMICAL.NAME) ])
mod.metabo.shrinked <- randomForest(t(r.metabo$E[sel, ]), response, importance=TRUE)
imp <- mod.metabo.shrinked$importance
sel <- rownames(imp)[ order(-imp[,4]) ][1:10]
mod.metabo.shrinked <- randomForest(t(r.metabo$E[sel, ]), response, importance=TRUE)

pred.metabo.models <- sapply(ss4, function(s) {
  ## first a full model
  y.resp <- set[ colnames(s), "group" ]
  xx <- xpredict(x=r.metabo$E, x.response=response, y=s, y.response=y.resp, return.list=TRUE, rf=mod.metabo.full)
  r  <- xx$ret
  ## now model shrinked to top 10 variables
  r2 <- xpredict(x=r.metabo$E, x.response=response, y=s, y.response=y.resp, return.list=TRUE, rf=mod.metabo.shrinked)
  list(full=xx, shrinked=r2)
}, simplify=F)
pred.metabo <- sapply(pred.metabo.models, function(m) list(full=m$full$ret, shrinked=m$shrinked$ret), simplify=FALSE)
pred.metabo <- unlist(pred.metabo, recursive=FALSE)

pred.metabo.tab <- rocseriestable(pred.metabo)

## model only on TOT
pred.metabo.tot <- xpredict(x=r.metabo$E, x.response=r.metabo$targets$group, y=ss4$TOT.all, y.response=set[ colnames(ss4$TOT.all), "group" ], return.list=TRUE)

## reverse predictions
pred.metabo.tot.rev <- xpredict(y=r.metabo$E, y.response=r.metabo$targets$group, x=ss4$TOT.all, x.response=set[ colnames(ss4$TOT.all), "group" ], return.list=TRUE)
```

```{r mlgambia}
## This data set is based on differential diagnosis between TB patients and
## patients suffering from other respiratory diseases and has been collected in The Gambia
set.seed(20180117)
sel <- Reduce(intersect, c(list(rownames(tbord.d)), sapply(ss4, rownames)))
response <- factor(tbord$group)
mod.jayne.full <- randomForest(t(tbord.d[sel,]), response, importance=TRUE)

pred.jayne <- sapply(ss4, function(s) {
      xpredict(x=tbord.d, x.response=tbord$group, y=s, y.response=set[ colnames(s), "group"], rf=mod.jayne.full)
}, simplify=F)

ord.rev.pred <- xpredict(x=ss3$TOT.all, x.response=set123[ colnames(ss3$TOT.all), "group" ], y=tbord.d, y.response=tbord$group)
pred.jayne.tot.model <- xpredict(y=ss3$TOT.all, y.response=set123[ colnames(ss3$TOT.all), "group" ], x=tbord.d, x.response=tbord$group, return.list=TRUE)
```

[//]: # (Similarity calculation between the Total, TB-HEALTHY and TB-ORD models)
```{r similarity}
load("data/rf.total.rda") # this is the final model saved prior to the unblinding
cor.models <- list()

imp <- list()
imp$metabo <- pred.metabo.tot$rf$importance[ order(pred.metabo.tot$rf$importance[,4], decreasing=T), ]
imp$jayne  <- pred.jayne.tot.model$rf$importance[ order(pred.jayne.tot.model$rf$importance[,4], decreasing=T), ]
imp$totalF <- rf.total$importance[ order(rf.total$importance[,4], decreasing=T), ]

sel <- intersect(rownames(imp$metabo), rownames(imp$totalF))
cor.models$metabo <- cor.test(imp$metabo[sel,4], imp$totalF[sel,4])
sel <- intersect(rownames(imp$jayne), rownames(imp$totalF))
cor.models$jayne  <- cor.test(imp$jayne[sel, 4], imp$totalF[sel,4])
```






[//]: # (Differential abundance analysis based on all data sets -----)
```{r diff,results="hide"}
## create a joint object containing metabolite measurements for all
## metabolites that can be found in all sample types
sel <- Reduce(intersect, lapply(list(serum, plasma, rpmi), rownames))
rg <- make.EList(data.matrix(cbind(serum[sel,], plasma[sel,], rpmi[sel,])), all.biochemicals, set)
rg$targets$group <- factor(rg$targets$group, levels=c("control", "case"))

## not interested in "unnamed" compounds
#rg <- rg[ !is.na(rg$genes$CHEMICAL.ID), ]

## rank normalization
rgR <- EList.ranknorm(rg)

## model 1: proximal
mod.1 <- model.fit(rgR,
  #selection=rgR$targets$revTP == "last",
  selection=rgR$targets$TIME_TO_TB < cutoff,
  formula="~ 0 + group + SITE + TIME_TO_TB + SEX + SAMPLE_TYPE + CASE_GROUP",
  contrasts=c(full="groupcase-groupcontrol"))

## model: sites
mod.sites <- sapply(sites, 
	function(s) 
		model.fit(rgR,
		selection=rgR$targets$SITE == s,
		formula="~ 0 + group + TIME_TO_TB + SEX + CASE_GROUP",
		contrasts=c(full="groupcase-groupcontrol")),
  simplify=FALSE)

#mod.trend <- list(trend=model.fit(
#  rgR,
#  selection=rgR$targets$group == "case",
#  formula="~ TIME_TO_TB + SITE + SEX + CASE_GROUP",
#  contrasts=NULL))

sets <- list(
  full=TRUE,
  proximate= rgR$targets$TIME_TO_TB < cutoff,
  distal=rgR$targets$TIME_TO_TB > cutoff
  )

mod.tps <- sapply(sets, function(s) 
  model.fit(rgR, selection=s, formula="~ 0 + group + SITE + SEX + SAMPLE_TYPE + CASE_GROUP", 
  contrasts=c(full="groupcase-groupcontrol")), 
  simplify=FALSE)

mods.all <- c(mod.tps, mod.sites)


## model 2: all time points
mod.2 <- model.fit(rgR,
  selection=TRUE,
  formula="~ 0 + group + SITE + TIME_TO_TB + SEX + SAMPLE_TYPE + CASE_GROUP",
  contrasts=c(full="groupcase-groupcontrol"))
```




[//]: # (TST: do we see a signature of conversion rather than incipient TB?)

```{r tst, results="hide"}
## select only clear conversion cases, and only in controls to eliminate
## signatures of disease
foo <- set[ set$TIMEPOINT != "BL" & set$TST >= 10, ]$DONOR_ID
foo <- set[ set$DONOR_ID %in% foo & set$TST < 10 & set$TST > -99 & set$TIMEPOINT == "BL" & set$CASE_CONTROL == "control", ]$DONOR_ID
foo1 <- set[ set$DONOR_ID %in% foo & set$TIMEPOINT == "BL", ]
foo2 <- set[ set$DONOR_ID %in% foo & set$TIMEPOINT != "BL" & set$TST > -99, ]
foo1 <- foo1[ match(foo, foo1$DONOR_ID), ]
foo2 <- foo2[ match(foo, foo2$DONOR_ID), ]
all(foo1$DONOR_ID == foo2$DONOR_ID) # TRUE
all(foo1$SAMPLE_ID %in% colnames(rg)) # TRUE

## paired wilcoxon test
foo1.smp <- match(foo1$SAMPLE_ID, colnames(rg))
foo2.smp <- match(foo2$SAMPLE_ID, colnames(rg))
foo.res <- apply(rg$E, 1, function(x) { wilcox.test(x[ foo1.smp ], x[ foo2.smp ], paired=T)$p.value })
sum(p.adjust(foo.res, method="fdr") < 0.05, na.rm=T) # 0
sum(p.adjust(foo.res, method="fdr") < 0.10, na.rm=T) # 0
sum(p.adjust(foo.res, method="fdr") < 0.20, na.rm=T) # 0

## test machine learning model to distinguish the after conversion samples
xx <- rg$E[ , c(foo1.smp, foo2.smp) ]
xx.g <- rep(c("BL", "conv"), each=length(foo1.smp))

rf <- randomForest(t(xx), factor(xx.g))
roc(response=rf$y, predictor=rf$votes[,1], ci=TRUE) # AUC 0.5625 95% CI: 0.2451-0.8799

## test in MRC only given the large number of TST-
## only controls, only baseline samples, only samples with determined test
## result
sel <- with(set, TIMEPOINT == "BL" & SITE == "MRC" & TST > -99 & CASE_CONTROL == "control")
xx <- t(rg$E[ , set$SAMPLE_ID[sel]])

## correlation between the TST reported size in mm and metabolic compounds
foo.pv <- apply(xx, 2, function(xxx) cor.test(xxx, set$TST[sel], method="s")$p.value)
sum(p.adjust(foo.pv, method="fdr") < 0.05, na.rm=T) # 0

## wilcoxon test to distinguish between TST- and TST+
foo.gr <- paste0("x", set$TST[sel] >= 10)
foo.pv <- apply(xx, 2, function(xxx) wilcox.test(xxx ~ foo.gr)$p.value)
sum(p.adjust(foo.pv, method="fdr") < 0.05, na.rm=T) # 0

## random forest. We don't crossvalidate, because we don't worry about
## overfitting - we worry only about lack of sensitivity.
set.seed(20180807)
rf <- randomForest(xx, factor(foo.gr))
roc(response=rf$y, predictor=rf$votes[,1], ci=TRUE) # AUC 0.5527 95% CI: 0.4393-0.6661

## same for SUN
## since the prevalence of TST positivity is different in different
## populations, one cannot bin the data
sel <- with(set, TIMEPOINT == "BL" & SITE == "SUN" & TST > -99 & CASE_CONTROL == "control")
xx <- t(rg$E[ , set$SAMPLE_ID[sel]])
foo.pv <- apply(xx, 2, function(xxx) cor.test(xxx, set$TST[sel], method="s")$p.value)
sum(p.adjust(foo.pv, method="fdr") < 0.01, na.rm=T) # 0

## wilcoxon test to distinguish between TST- and TST+
foo.gr <- paste0("x", set$TST[sel] >= 10)
foo.pv <- apply(xx, 2, function(xxx) wilcox.test(xxx ~ foo.gr)$p.value)
sum(p.adjust(foo.pv, method="fdr") < 0.05, na.rm=T) # 0

## RF
set.seed(20180807)
rf <- randomForest(xx, factor(foo.gr))
roc(response=rf$y, predictor=rf$votes[,1], ci=TRUE) # AUC 0.5717 95% CI: 0.4149-0.7284


## MAK. There are only TST+ controls in MAK, therefore only the correlation
## is tested
sel <- with(set, TIMEPOINT == "BL" & SITE == "MAK" & TST > -99 & CASE_CONTROL == "control")
xx <- t(rg$E[ , set$SAMPLE_ID[sel]])
foo.pv <- apply(xx, 2, function(xxx) cor.test(xxx, set$TST[sel], method="s")$p.value)
sum(p.adjust(foo.pv, method="fdr") < 0.01, na.rm=T) # 0

## same for AHRI
sel <- with(set, TIMEPOINT == "BL" & SITE == "AHRI" & TST > -99 & CASE_CONTROL == "control")
xx <- t(rg$E[ , set$SAMPLE_ID[sel]])
foo.pv <- apply(xx, 2, function(xxx) cor.test(xxx, set$TST[sel], method="s")$p.value)
sum(p.adjust(foo.pv, method="fdr") < 0.01, na.rm=T) # 0

## wilcoxon test to distinguish between TST- and TST+
foo.gr <- paste0("x", set$TST[sel] >= 10)
foo.pv <- apply(xx, 2, function(xxx) wilcox.test(xxx ~ foo.gr)$p.value)
sum(p.adjust(foo.pv, method="fdr") < 0.05, na.rm=T) # 0

## RF
set.seed(20180807)
tst.ahri.rf <- randomForest(xx, factor(foo.gr))
roc(response=tst.ahri.rf$y, predictor=tst.ahri.rf$votes[,1], ci=TRUE) # AUC 0.8968 95% CI: 0.7852-1 [!!!]
tst.ahri.gr <- foo.gr
tst.ahri.pred <- data.frame(prediction=tst.ahri.rf$votes[,1], reality=tst.ahri.gr)
tst.ahri.tab <- rocseriestable(list(a=tst.ahri.pred), levels=c("xFALSE", "xTRUE"))

## 

```






[//]: # (Sample type specific differential abundance analysis -----)
```{r diff_spec,results="hide"}
## plasma, serum and rpmi specific models
## only for compounds not in the general models to save on statistical
## power
mod.plasma <- model.fit(r.plasmaR,
  selection=TRUE,
  formula="~ 0 + group + TIME_TO_TB + SEX + CASE_GROUP",
  contrasts=c(full="groupcase-groupcontrol"))

mod.serum <- model.fit(r.serumR,
  selection=TRUE,
  formula="~ 0 + group + TIME_TO_TB + SEX + CASE_GROUP",
  contrasts=c(full="groupcase-groupcontrol"))

mod.rpmi <- model.fit(r.rpmiR,
  selection=TRUE,
  formula="~ 0 + group + TIME_TO_TB + SEX + CASE_GROUP",
  contrasts=c(full="groupcase-groupcontrol"))
```


[//]: # (Joint results table for all models -----)
```{r diff_jointTT,results="hide"}

## a joint results table for the different models
all.models <- list(mod.1, mod.2, mod.plasma, mod.serum, mod.rpmi)
names(all.models) <- c("mod.1", "mod.2", "mod.plasma", "mod.serum", "mod.rpmi")

tts <- lapply(all.models, 
  function(x) topTable(x$fit, number=Inf, sort.by="n")[,c("ID", "adj.P.Val")])
names(tts) <- c("mod.1", "mod.2", "mod.plasma", "mod.serum", "mod.rpmi")
jointTT <- Reduce(function(x,y) merge(x, y, by.x="ID", by.y="ID", all=T), tts)
colnames(jointTT)[-1] <- names(tts)
jointTT$min.P <- apply(jointTT[,c("mod.1", "mod.2")], 1, function(x) min(x, na.rm=T))
jointTT <- jointTT[order(jointTT$min.P), ]
jointTT <- data.frame(ID=jointTT$ID, Name=all.biochemicals[ jointTT$ID, "BIOCHEMICAL" ],
                      Group=all.biochemicals[ jointTT$ID, "SUPER.PATHWAY" ],
                      jointTT[,-1])
nn.joint <- apply(jointTT[,names(all.models)], 2, function(x) sum(x < 0.05, na.rm=T))

## mod contains residual variation from a stratification model for making nice plots
mod.resid <- lm(t(rg$E) ~ SITE + SAMPLE_TYPE + CASE_GROUP + SEX, data=rgR$targets)
mod.resid <- t(mod.resid$residuals)
sel <- rgR$targets$group == "control"
#mod <- t(apply(rgR$E, 1, function(x) { (x - median(x[sel])) / IQR(x[sel]) }))
mod <- t(apply(rgR$E, 1, scale))
mod <- t(apply(rgR$E, 1, function(x) { (x - median(x[sel])) / IQR(x[sel]) }))
```

[//]: # (glm models - binomial logit regression -----)
```{r glm,cache=TRUE}
if(file.exists("static_cache/glm_models.rgl")) {
  load("static_cache/glm_models.rgl") 
} else {
  formula1 <- "group ~ x + SAMPLE_TYPE + CASE_GROUP"
  formula2 <- "group ~     SAMPLE_TYPE + CASE_GROUP"
  glm.full      <- run.glm(rg, mod1=formula1, mod2=formula2)
  glm.proximate <- run.glm(rg[ , rg$targets$TIME_TO_TB <  cutoff | rg$targets$group == "control" ], mod1=formula1, mod2=formula2)
  glm.distal    <- run.glm(rg[ , rg$targets$TIME_TO_TB >= cutoff | rg$targets$group == "control" ], mod1=formula1, mod2=formula2)
  save(glm.full, glm.proximate, glm.distal, file="static_cache/glm_models.rgl")
}
```

[//]: # (Time dependence models)
```{r diff_time,cache=TRUE}
message("Running diff_time")
require(lme4)
sel <- rg$targets$group == "case"
rgC <- rg[,sel]

y <- t(rgC$E)
lm.mod <- summary(lm(y ~ TIME_TO_TB + SAMPLE_TYPE + SEX + DONOR_ID, data=rgC$targets))
lm.mod.res <- sapply(lm.mod, function(x) x$coefficients["TIME_TO_TB", "Pr(>|t|)"])
names(lm.mod.res) <- gsub(".* ", "", names(lm.mod.res))

if(file.exists("static_cache/lmer.res.rda")) {
  load("static_cache/lmer.res.rda")
} else {
  mod1 <- "y ~              SAMPLE_TYPE + CASE_GROUP + (1|DONOR_ID)"
  mod2 <- "y ~ TIME_TO_TB + SAMPLE_TYPE + CASE_GROUP + (1|DONOR_ID)"
  lmer.res <- run.lmer.2(rgC, mod1, mod2)
  save(lmer.res, file="static_cache/lmer.res.rda")
}
## we do the same using a mixed effects model
##   message("Running lmer.0")
##   lmer.0 <- apply(rgC$E, 1, function(y) lmer(y ~ SAMPLE_TYPE + SEX + (1|DONOR_ID), data=rgC$targets, REML=F))
##   message("Running lmer.M")
##   lmer.M <- apply(rgC$E, 1, function(y) lmer(y ~ TIME_TO_TB + SAMPLE_TYPE + SEX + (1|DONOR_ID), data=rgC$targets, REML=F))
##   message("done")
##   lmer.res <- t(sapply(names(lmer.0), function(n) { xx <- anova(lmer.0[[n]], lmer.M[[n]]) ; Reduce(rbind, xx) }))
##   colnames(lmer.res) <- c("df.1", "AIC.1", "BIC.1", "LL.1", "DEV.1", "x", "x1", "x2", "df.2", "AIC.2", "BIC.2", "LL.2", "DEV.2", "Chisq", "Chi_df", "p")
##   lmer.res <- lmer.res[ , -c(6:8) ]
##   lmer.res <- data.frame(rgC$genes[ , c("ID", "BIOCHEMICAL", "HMDB")], lmer.res[,c("AIC.1", "AIC.2", "p")])
##   lmer.res$p.adj <- p.adjust(lmer.res$p, method="fdr")
##   #lmer.res <- run.lmer(rgC, "y ~ SAMPLE_TYPE + SEX + (1|DONOR_ID)", "y ~ TIME_TO_TB + SAMPLE_TYPE + SEX + (1|DONOR_ID)")
##   lmer.res$lm.mod.p <- lm.mod.res
##   lmer.res$lm.mod.p.adj <- p.adjust(lmer.res$lm.mod.p, method="fdr")


# enrichment test using tmod
lmer.enr.res <- tmodUtest(rgC$genes$HMDB[ order(lmer.res$P.Value) ], mset=mm, qval=1)
# only positive enrichment
lmer.enr.res <- lmer.enr.res[ lmer.enr.res$AUC > 0.5, ]
lmer.enr.res$adj.P.Val <- p.adjust(lmer.enr.res$P.Value, method="fdr")
```

[//]: # (Metabolite classification based on the modeling results)

```{r classification,cache=TRUE}
ids <- rownames(glm.full)
classification <- data.frame(ID=ids,
  Description=all.biochemicals[ids, "BIOCHEMICAL"],
  GLM.D=glm.distal[ids, ]$adj.P.Val,
  GLM.P=glm.proximate[ids, ]$adj.P.Val,
  GLM.F=glm.full[ids, ]$adj.P.Val,
  LMER=lmer.res[ids, ]$adj.P.Val)
rownames(classification) <- ids

classification$class <- "" # undefined

sel.dis <- with(classification, 
  GLM.D > 0.05 & (LMER < 0.05 | GLM.P < 0.05))

sel.risk <- with(classification,
  GLM.D < 0.01 & GLM.P > 0.01 & (LMER > 0.1 | (LMER < 0.05 & GLM.P > 0.1)))

classification$class[sel.risk] <- "risk-associated"
classification$class[sel.dis] <- "disease-associated"
```





```{r smoking}
## SMOKING ISSUE

# order of smoking levels to have "No smoking" as reference

t.SMOKE <- factor(rg$targets$SMOKE, levels=c("No", "> 6M", "< 6M", "Yes"))
t.group <- factor(rg$targets$group, levels=c("control", "case"))

# all data
smoke.m  <- glm(t.group ~ t.SMOKE, family=binomial(logit))

smoke.coef <- summary(smoke.m)$coefficients[c("t.SMOKE> 6M","t.SMOKE< 6M","t.SMOKEYes"),]
smoke.p    <- 1 - pchisq(smoke.m$null.deviance - smoke.m$deviance, smoke.m$df.null - smoke.m$df.residual)
smoke.or   <- exp(smoke.coef[,1])

smoke.ci.l <- smoke.or / exp(qnorm(0.975) * smoke.coef[,2])
smoke.ci.r <- smoke.or * exp(qnorm(0.975) * smoke.coef[,2])

res.all.smoke <- cbind(smoke.or, smoke.ci.l, smoke.ci.r, smoke.coef[,4], smoke.p)
colnames(res.all.smoke ) <- c("OR", "CI.L", "CI.R", "Beta.P.Value","LRT")

# by SITE
t.site <- setdiff(sites, "AHRI") # no available data

smoke.site <- sapply(t.site, function(site) {
  sel <- rg$targets$SITE == site
  t.s <- t.SMOKE[ sel ]
  t.g <- t.group[ sel ]
	m  <- glm(t.g ~ t.s, family=binomial(logit))
  name <- rownames(summary(m)$coefficients)
  coef <- summary(m)$coefficients[name[2:length(name)],]
  p <- 1 - pchisq(m$null.deviance - m$deviance, m$df.null - m$df.residual)
  or   <- exp(coef[,1])
  ci.l <- or / exp(qnorm(0.975) * coef[,2])
  ci.r <- or * exp(qnorm(0.975) * coef[,2])
  data.frame(or=or, ci.l=ci.l, ci.r=ci.r, beta=coef[,4], P.Value=p)
}, simplify=FALSE)

```





